Schema Evolution in Interactive Programming Systems

### 2024-08-24 paper.tex
typos

In this paper we identify a common problem, \emph{Schema Change}, that stands athwart feedback loops in diverse contexts of software development. Schema Change has been extensively studied in the context of databases where it necesitates manual and error-prone data migrations.~\cite{erhard06} We observe analogous problems occuring in varied and less-studied contexts. For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't. Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system. We discuss further examples in \S\ref{sec:related} Related Work. In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops. To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also data types and specifications of all sorts, whether expressed in a formal language or left implicit in code.

Our primary contribution is a suite of challenge problems exemplifying a diverse range of typical schema change scenarios that interrupt feedback loops. We believe these problems have not been completely solved in theory or practice, which we discuss further in \S\ref{sec:related} Related Work. While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches. None of our ongoing efforts yet address the full range of problems. Indeed the motivation of this paper is to provide a basis for communicating and collaborating between such related efforts, as well as to facilitate further research efforts. We hope these challange problems serve as examples to ground future research and as a basis for evaluating its results. We invite contributions of new challenge problems and look forward to discussing competing solutions.

@@

The schema and data evolution happens through a sequence of edits. Table~\ref{tbl:docchanges}
lists edits appearing in our examples (but not a complete list). A system may
use a structure editor that provides high-level commands for triggering those edits.
Recognising user intent from plain text editing of source code may make the challenge harder.

@@

This challenge envisions a new feedback channel that end users have not experienced before: the ability to frictionlessly exchange variations and elaborations they make to docuuments, replacing the hellscape of manual copy \& paste. But meeting this challenge may require a basic change of perspective in both live programming and local-first software.
Live programming must move from being a solitary activity to a collaborative workflow, and further where the collaboration is not just on editing source code but on live integrated code and data,
as in the Smalltalk/Lisp images of old.
%\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!}
Local-first software must move from automatically converging data replicas
to also interactively managing long-term divergence. That implies either application state
is being directly exposed to the user, or the application itself offers
version control affordances to the user. We encourage both communities to think outside the box of Git,
which has proven to baffle not only end-users but also a substantial fraction of developers. The Divergence challenge may be as much a technical challenge as a conceptual challenge, asking us to reconsider the entrenched boundaries between technologies in the stack.

@@

In the following we assume that the state machine is executed by an interpreter that processes
incoming events, traverses the transitions accordingly, and executes on-entry actions. The notion
of the current state and the values of the global variables are the collective run-time state of
the program. The structure that the interpreter operates on can be described by a run-time schema.
This schema (Figure~\ref{FIG:doorsRuntime}, left) is derived from the abstract syntax
schema (Figure~\ref{FIG:statemachineUML}, left), but is decorated with additional properties and
associations modeling the run-time state required for execution. The augmented metamodel includes
properties \lstinline{vars} (to model variable assigments) and \lstinline{visited} (counting
number of state visits) to the classes \lstinline{Machine} and \lstinline{State}, respectively.

@@

\paragraph{Schema evolution.}
Schema evolution has long been a major problem for SQL databases because
the SQL Data Definition Language (DDL) has limited capabilties to redefine existing data. It can rename tables and columns. SQLite \cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
MYSql \cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for data migration. As a result in practice schema evolution is mostly done by writing custom Data Manipulation Language (DML) code to migrate the data. Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
In \emph{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.

@@

Some tools provide a Domain Specific Language (DSL) to describe schema evolution. Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to manage schema migration but is comparable to the capabilities of SQL DDL, often requiring the addition of custom Ruby or SQL code. Schema evolution is related to the classical view-update problem~\cite{Bancilhon81} which was approached from a language perspective by lenses~\cite{Foster2007}. Generalizing from lenses \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''. SMOs can also rewrite queries to accomodate schema changes. \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}. BiDEL appears capable of handling the basic requirements of our \emph{Extract/Absorb Entity} and \emph{Multiplicity Change} challenges, though \emph{Split/Merge Entity} is less clear. It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.

@@

\paragraph{Model Driven Engineering.}
Model Driven Engineering must also handle evolution. Unlike the textual artifacts involved in other domains, models typically assign unique indentifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
Models are often themselves modeled by a metamodel, which lifts the evolution problem to the metalevel: models must be migrated through the evolution of their metamodel, analogously to changing the grammar of a programming language.~\cite{Herrmannsdoerfer11}. \citet{Cicchetti11} study metamodel evolution in the context of divergent changes.
\citet{vermolen11} define a DSL for evolving a data model by generating SQL migrations on the backing database. They consider the evolution of OO-style subclass hierarchies which goes beyond most other work. Their running example includes both our \emph{Extract Entity} and \emph{Multiplicty Change} examples, and would make for a good follow-on problem. However they do not address code/query rewriting.

>>>>

In this paper we identify a common problem, \emph{Schema Change}, that stands athwart feedback loops in diverse contexts of software development. Schema Change has been extensively studied in the context of databases where it necessitates manual and error-prone data migrations.~\cite{erhard06} We observe analogous problems occuring in varied and less-studied contexts. For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't. Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system. We discuss further examples in \S\ref{sec:related} Related Work. In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops. To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also data types and specifications of all sorts, whether expressed in a formal language or left implicit in code.

Our primary contribution is a suite of challenge problems exemplifying a diverse range of typical schema change scenarios that interrupt feedback loops. We believe these problems have not been completely solved in theory or practice, which we discuss further in \S\ref{sec:related} Related Work. While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches. None of our ongoing efforts yet address the full range of problems. Indeed the motivation of this paper is to provide a basis for communicating and collaborating between such related efforts, as well as to facilitate further research efforts. We hope these challenge problems serve as examples to ground future research and as a basis for evaluating its results. We invite contributions of new challenge problems and look forward to discussing competing solutions.

@@

The schema and data evolution happens through a sequence of edits. Table~\ref{tbl:docchanges}
lists edits appearing in our examples (but not a complete list). A system may
use a structure editor that provides high-level commands for triggering those edits.
Recognizing user intent from plain text editing of source code may make the challenge harder.

@@

This challenge envisions a new feedback channel that end users have not experienced before: the ability to frictionlessly exchange variations and elaborations they make to documents, replacing the hellscape of manual copy \& paste. But meeting this challenge may require a basic change of perspective in both live programming and local-first software.
Live programming must move from being a solitary activity to a collaborative workflow, and further where the collaboration is not just on editing source code but on live integrated code and data,
as in the Smalltalk/Lisp images of old.
%\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!}
Local-first software must move from automatically converging data replicas
to also interactively managing long-term divergence. That implies either application state
is being directly exposed to the user, or the application itself offers
version control affordances to the user. We encourage both communities to think outside the box of Git,
which has proven to baffle not only end-users but also a substantial fraction of developers. The Divergence challenge may be as much a technical challenge as a conceptual challenge, asking us to reconsider the entrenched boundaries between technologies in the stack.

@@

In the following we assume that the state machine is executed by an interpreter that processes
incoming events, traverses the transitions accordingly, and executes on-entry actions. The notion
of the current state and the values of the global variables are the collective run-time state of
the program. The structure that the interpreter operates on can be described by a run-time schema.
This schema (Figure~\ref{FIG:doorsRuntime}, left) is derived from the abstract syntax
schema (Figure~\ref{FIG:statemachineUML}, left), but is decorated with additional properties and
associations modeling the run-time state required for execution. The augmented metamodel includes
properties \lstinline{vars} (to model variable assignments) and \lstinline{visited} (counting
number of state visits) to the classes \lstinline{Machine} and \lstinline{State}, respectively.

@@

\paragraph{Schema evolution.}
Schema evolution has long been a major problem for SQL databases because
the SQL Data Definition Language (DDL) has limited capabilities to redefine existing data. It can rename tables and columns. SQLite \cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
MYSql \cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for data migration. As a result in practice schema evolution is mostly done by writing custom Data Manipulation Language (DML) code to migrate the data. Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
In \emph{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.

@@

Some tools provide a Domain Specific Language (DSL) to describe schema evolution. Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to manage schema migration but is comparable to the capabilities of SQL DDL, often requiring the addition of custom Ruby or SQL code. Schema evolution is related to the classical view-update problem~\cite{Bancilhon81} which was approached from a language perspective by lenses~\cite{Foster2007}. Generalizing from lenses \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''. SMOs can also rewrite queries to accommodate schema changes. \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}. BiDEL appears capable of handling the basic requirements of our \emph{Extract/Absorb Entity} and \emph{Multiplicity Change} challenges, though \emph{Split/Merge Entity} is less clear. It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.

@@

\paragraph{Model Driven Engineering.}
Model Driven Engineering must also handle evolution. Unlike the textual artifacts involved in other domains, models typically assign unique identifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
Models are often themselves modeled by a metamodel, which lifts the evolution problem to the metalevel: models must be migrated through the evolution of their metamodel, analogously to changing the grammar of a programming language.~\cite{Herrmannsdoerfer11}. \citet{Cicchetti11} study metamodel evolution in the context of divergent changes.
\citet{vermolen11} define a DSL for evolving a data model by generating SQL migrations on the backing database. They consider the evolution of OO-style subclass hierarchies which goes beyond most other work. Their running example includes both our \emph{Extract Entity} and \emph{Multiplicity Change} examples, and would make for a good follow-on problem. However they do not address code/query rewriting.

### 2024-08-24 revisions.tex
typos

We thank the anonymous reviewers for their helpful comments. In the following we summarize the changes we have made in response to those comments.

@@

In the introduction we clarify that this paper is not about a tool under development. Some of the problems are extracted from our different research projects but none of these projects span the full range of problems.

We have clarified in the introduction that we avoid speculating about possible solutions to the challenge problems so as to not bias them towards our preferred approaches (and to avoid the issues where we disagree). A primary goal of this paper is to establish a neutral common ground where researchers can meet to discuss their alternative approaches.
\textcolor{red}{Or do we want to add a speculative section?}

@@

We rewrote the introduction to clarify that the goal of our research on schema change is to speed feedback loops in software development. Ultimately that is a value judgement, but we believe that value is important to a substantial community of researchers and practitioners. We have qualified statements of opinion throughout the paper.

Section 2 has been revised to remove the claim of providing a conceptual framework. Instead we define a descriptive framework of multiple dimensions and layers of schema change that are used to figuratively contrast the different problems.

\textcolor{red}{I don't know what to do with this:}
\begin{displayquote}
  Just before 2.2 I think here you need to set expectations for correct local transformations, because no general solution exists, only conventions and expectations. Please, add numbers to those expectations, show examples of corner cases where stuff breaks and is an acceptable evil, and cases were stuff breaks on purpose (in some sense, and intended evil) To make an example: Java is of course not truly "cross-platform": if you wanted to put a formal definition to cross platform, the program should work identically in all environments. This would imply that the program should not be able to discover it's running environment. In Java I can make an 'if' at the beginning of the main and throw error on Windows. This makes Java not truly "cross-platform" in a logical sense, but this is often considered desirable.

@@

We clarified that the goal of Divergence in End-User Workflows is to create a new kind of feedback channel for end users rather than speeding up existing channels.

In \S\emph{Remarks: Managing Design Tradeoffs} we noted the importance of offering solutions simple enough to be actually used.

\section{Reviewer B}

As suggested we have replaced `type' with `schema' throughout the paper. Some diagrams still use the letter `T' for type but we intend to change those to `S' in the camera-ready version.

\textcolor{red}{I don't know what to do with this:}
\begin{displayquote}
I liked the discussion of both local/singular and collaborative development and the problems associated with them. And I would have liked to see those two situations discussed separately, because of their technical complexity and also the processes (people) causing them.

@@

\textcolor{red}{I don't know what to do with this:}
\begin{displayquote}
  While re-reading the submission, I noticed quite a similarity to a pattern language (in the sense of Christopher Alexander and/or the software patterns community). All sections covering a challenge problem present the context in which the problems occurs, describe the problem, and provide an example illustrating the problem to better relate to it. Also, they cross-reference other related challenge problems. The only obvious part that is missing (on purpose, see above) is a generative solution.

@@

As suggested, the Literature Review was converted into a more appropriately named Related Work section.

\section{Reviewer C}

in \S\emph{Remarks: Requirements and Implementation} we clarify that the goal is to adapt to schema changes automatically to avoid interrupting live programming.

We added to Related Work the history of how the view-update problem led to lenses which led to DELs.

We cited Hazel in Related Work.

We now mention the possibility of using CRDTs for evolving computational documents.

Divergence Control for Spreadsheets has been corrected to just Divergence Control.

As mentioned above, we have added an explanation in the introduction of why we maintain neutrality by avoiding speculation on our preferred solutions.

\end{document}

>>>>

We thank the anonymous reviewers for their helpful comments. In the following we summarize the changes we have made in response.

@@

\begin{itemize}
\item In the introduction we clarify that this paper is not about a tool under development. Some of the problems are extracted from our different research projects but none of these projects span the full range of problems.

\item We have clarified in the introduction that we avoid speculating about possible solutions to the challenge problems so as to not bias them towards our preferred approaches (and to avoid the issues where we disagree). A primary goal of this paper is to establish a neutral common ground where researchers can meet to discuss their alternative approaches.
\textcolor{red}{Or do we want to add a speculative section?}
\end{itemize}

@@

\begin{itemize}
  \item We rewrote the introduction to clarify that the goal of our research on schema change is to speed feedback loops in software development. Ultimately that is a value judgement, but we believe that value is important to a substantial community of researchers and practitioners. We have qualified statements of opinion throughout the paper.

  \item Section 2 has been revised to remove the claim of providing a conceptual framework. Instead we define a descriptive framework of multiple dimensions and layers of schema change that are used to figuratively contrast the different problems.

  \item \textcolor{red}{I don't know what to do with this:}
\begin{displayquote}
  Just before 2.2 I think here you need to set expectations for correct local transformations, because no general solution exists, only conventions and expectations. Please, add numbers to those expectations, show examples of corner cases where stuff breaks and is an acceptable evil, and cases were stuff breaks on purpose (in some sense, and intended evil) To make an example: Java is of course not truly "cross-platform": if you wanted to put a formal definition to cross platform, the program should work identically in all environments. This would imply that the program should not be able to discover it's running environment. In Java I can make an 'if' at the beginning of the main and throw error on Windows. This makes Java not truly "cross-platform" in a logical sense, but this is often considered desirable.

@@

\item We clarified that the goal of Divergence in End-User Workflows is to create a new kind of feedback channel for end users rather than speeding up existing channels.

\item In \S\emph{Remarks: Managing Design Tradeoffs} we noted the importance of offering solutions simple enough to be actually used.
\end{itemize}

\section{Reviewer B}
\begin{itemize}

  \item As suggested we have replaced `type' with `schema' throughout the paper. Some diagrams still use the letter `T' for type but we intend to change those to `S' in the camera-ready version.

  \item \textcolor{red}{I don't know what to do with this:}
\begin{displayquote}
I liked the discussion of both local/singular and collaborative development and the problems associated with them. And I would have liked to see those two situations discussed separately, because of their technical complexity and also the processes (people) causing them.

@@

\item \textcolor{red}{I don't know what to do with this:}
\begin{displayquote}
  While re-reading the submission, I noticed quite a similarity to a pattern language (in the sense of Christopher Alexander and/or the software patterns community). All sections covering a challenge problem present the context in which the problems occurs, describe the problem, and provide an example illustrating the problem to better relate to it. Also, they cross-reference other related challenge problems. The only obvious part that is missing (on purpose, see above) is a generative solution.

@@

\item As suggested, the Literature Review was converted into a more appropriately named Related Work section.
\end{itemize}

\section{Reviewer C}
\begin{itemize}
  \item in \S\emph{Remarks: Requirements and Implementation} we clarify that the goal is to adapt to schema changes automatically to avoid interrupting live programming.

  \item Added to Related Work the history of how the view-update problem led to lenses which led to DELs.

  \item Cited Hazel in Related Work.

  \item Mention the possibility of using CRDTs for evolving computational documents.

  \item Divergence Control for Spreadsheets has been corrected to just Divergence Control.

  \item As mentioned above, we have added an explanation in the introduction of why we maintain neutrality by avoiding speculation on our preferred solutions.
\end{itemize}
\end{document}
